---
title: "R Code Summary"
format: html
editor: visual
---

## Data visualization

### `ggplot(data, aes(...)) + geom_*()`

```{r}
library(ggplot2)

ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

-   Maps variables to plot aesthetics (here: `displ` (engine displacement)→ x, `hwy` (highway miles per gallon)→ y, `class` → color).
-   Draws a layer (here points) using a `geom_*()` function.

### `facet_wrap(~var)`

```{r}
library(ggplot2)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~ class)
```

-   Splits one plot into multiple panels by a categorical variable.
-   Lets you compare patterns across groups on the same axes/scales.

### `ggsave("file.png")`

```{r}
library(ggplot2)

p <- ggplot(mpg, aes(displ, hwy)) + geom_point()
ggsave("mpg_scatter.png", plot = p, width = 6, height = 4)
```

-   Saves a plot object (or last plot) to a file.
-   Controls output size via `width` and `height`.

## Spatial data

### `rnaturalearth::ne_countries(..., returnclass="sf")`

```{r}
library(sf)
library(rnaturalearth)

world <- ne_countries(scale = "medium", returnclass = "sf")
```

-   Downloads country boundaries from Natural Earth.
-   Returns an `sf` object (geometry + attribute columns).

### `ggplot(sf_obj) + geom_sf()`

```{r}
library(ggplot2)

ggplot(world) +
  geom_sf()
```

-   Plots spatial geometries stored in an `sf` object.
-   Automatically uses the geometry column for map drawing.

### `coord_sf(crs = "...")`

```{r}
library(ggplot2)

ggplot(world) +
  geom_sf() +
  coord_sf(crs = "EPSG:3857")
```

-   Reprojects spatial data to a chosen coordinate reference system (CRS).
-   Changes the map’s appearance/distortion based on the projection.

## Logical vectors

### `if_else(condition, true, false)`

```{r}
library(dplyr)

tibble(x = c(-2, 0, 5)) |>
  mutate(sign = if_else(x >= 0, "nonnegative", "negative"))
```

-   Creates a new vector by choosing between two values based on a condition.
-   Is type-stable (returns consistent type across all rows).

### `case_when(...)`

```{r}
library(dplyr)

tibble(score = c(95, 82, 71, 60)) |>
  mutate(letter = case_when(
    score >= 90 ~ "A",
    score >= 80 ~ "B",
    score >= 70 ~ "C",
    TRUE        ~ "D"
  ))
```

-   Applies multiple conditions in order; first match is used.
-   Produces a single output column from many branching rules.

### `is.na(x)`

```{r}
x <- c(1, NA, 3, NA)
is.na(x)
```

-   Returns a logical vector indicating which elements are missing.
-   Often used for filtering, counting, or conditional replacement.

## Numbers

### `count(x, wt = w)`

```{r}
library(dplyr)

tibble(team = c("A","A","B","B"), points = c(3, 2, 5, 1)) |>
  count(team, wt = points, name = "total_points")
```

-   Groups by a variable and sums weights (`wt`) instead of counting rows.
-   Produces a compact frequency/total table.

### `%/%` and `%%` (integer division + remainder)

```{r}
library(dplyr)

time <- c(730, 945, 1205)  # HHMM format as numbers
hour <- time %/% 100
minute <- time %% 100
tibble(time, hour, minute)
```

-   `%/%` divides and drops the remainder (integer division).
-   `%%` returns the remainder (useful for splitting encoded numbers).

### `lead()` / `lag()`

```{r}
library(dplyr)

tibble(day = 1:5, value = c(10, 11, 15, 14, 20)) |>
  mutate(change_from_prev = value - lag(value))
```

-   `lag()` shifts values down to compare with the previous row.
-   `lead()` shifts values up to compare with the next row.

### `log2()` / `log10()`

```{r}
x <- c(1, 10, 100, 1000)
log10(x)
```

-   Transforms multiplicative scale data into additive scale.
-   Helps compress large ranges and reveal proportional changes.

## Strings

### `str_c(...)`

```{r}
library(stringr)

first <- c("Jack", "Ada")
last  <- c("Dong", "Lovelace")

str_c(first, " ", last)
```

-   Concatenates strings element-by-element (vectorized).
-   Lets you control separators via explicit `" "` or `sep=` patterns.

### `str_glue("...{var}...")`

```{r}
library(glue)

name <- "Jack"
score <- 98
str_glue("{name} scored {score} on the exam.")
```

-   Builds strings using `{}` placeholders for variables/expressions.
-   Makes readable “template” text for labels, messages, filenames, etc.

### `str_length()` and `str_sub()`

```{r}
library(stringr)

s <- c("Macalester", "psychology")
str_length(s)
str_sub(s, 1, 3)
```

-   `str_length()` counts characters in each string.
-   `str_sub()` extracts substring by start/end positions.

## Regular expressions

### `str_detect(x, pattern)`

```{r}
library(stringr)

emails <- c("a@macalester.edu", "b@gmail.com", "c@macalester.edu")
str_detect(emails, "@macalester\\.edu$")
```

-   Returns TRUE/FALSE for whether each string matches the regex.
-   Useful for filtering rows that contain a pattern.

### `str_extract(x, pattern)`

```{r}
library(stringr)

texts <- c("Order #A123", "Order #B55", "No order")
str_extract(texts, "#[A-Z][0-9]+")
```

-   Extracts the first matching substring from each element.
-   Returns `NA` when no match exists.

### `str_replace()` / `str_replace_all()`

```{r}
library(stringr)

x <- c("apple pie", "apple juice")
str_replace(x, "apple", "orange")
str_replace_all("a1 b2 c3", "[0-9]", "X")
```

-   Replaces first match (`str_replace`) or all matches (`str_replace_all`).
-   Great for cleaning text, standardizing formats, removing junk characters.

## Factors

### `fct_reorder(f, x)`

```{r}
library(dplyr)
library(forcats)
library(ggplot2)

df <- tibble(
  group = factor(c("A","A","B","B","C","C")),
  score = c(5, 6, 2, 3, 9, 8)
)

df |>
  mutate(group = fct_reorder(group, score, .fun = mean)) |>
  ggplot(aes(group, score)) +
  geom_boxplot()
```

-   Reorders factor levels using a numeric summary (here mean score).
-   Makes plots and tables follow a meaningful order.

### `fct_relevel(f, "level")`

```{r}
library(forcats)

f <- factor(c("medium","low","high","low"))
fct_relevel(f, "low", "medium", "high")
```

-   Manually sets the order of factor levels.
-   Useful when you want a specific “baseline” or display order.

### `fct_infreq()` and `fct_rev()`

```{r}
library(forcats)

f <- factor(c("cat","dog","dog","dog","bird"))
fct_infreq(f)
fct_rev(fct_infreq(f))
```

-   `fct_infreq()` orders levels from most frequent to least frequent.
-   `fct_rev()` reverses the level order (often for plotting).

### `fct_recode()` / `fct_collapse()`

```{r}
library(forcats)

f <- factor(c("freshman","sophomore","junior","senior","senior"))

fct_recode(f, "1st" = "freshman", "2nd" = "sophomore")

fct_collapse(
  f,
  "underclass" = c("freshman","sophomore"),
  "upperclass" = c("junior","senior")
)
```

-   `fct_recode()` renames levels without changing grouping.
-   `fct_collapse()` merges several levels into fewer categories.

## Dates and times

### `today()` / `now()` and `ymd()` / `ymd_hms()`

```{r}
library(lubridate)

today()
now()

ymd("2026-02-20")
ymd_hms("2026-02-20 14:30:00")
```

-   Gets current date/time (`today`, `now`) or parses strings into date objects.
-   Converts text into proper date/datetime types for math and plotting.

### `make_date()` / `make_datetime()`

```{r}
library(lubridate)

make_date(2026, 2, 20)
make_datetime(2026, 2, 20, 14, 30, 0)
```

-   Builds date/datetime from separate components.
-   Helpful when your data has year/month/day in separate columns.

### `floor_date()` / `ceiling_date()`

```{r}
library(lubridate)

dt <- ymd_hms("2026-02-20 14:37:22")
floor_date(dt, "hour")
ceiling_date(dt, "hour")
```

-   Rounds date-times down/up to a specified unit (hour/day/week/etc.).
-   Useful for aggregating by time bins.

### `with_tz()` vs `force_tz()`

```{r}
library(lubridate)

dt <- ymd_hms("2026-02-20 12:00:00", tz = "UTC")

with_tz(dt, "America/Chicago")
force_tz(dt, "America/Chicago")
```

-   `with_tz()` changes how the same instant is displayed in another time zone.
-   `force_tz()` keeps clock time but changes the assumed time zone (different instant).

## Missing values

### `tidyr::fill(cols)`

```{r}
library(tidyr)
library(dplyr)

df <- tibble(group = c("A", NA, NA, "B", NA), value = 1:5)

df |>
  fill(group)
```

-   Fills missing values down (or up) using the nearest non-missing value.
-   Useful for “merged-cell” style data.

### `coalesce(x, 0)`

```{r}
library(dplyr)

x <- c(1, NA, 3)
coalesce(x, 0)
```

-   Replaces `NA` with the first non-missing value among arguments.
-   Great for setting defaults (e.g., missing counts become 0).

### `na_if(x, sentinel)`

```{r}
library(dplyr)

x <- c(10, -99, 7, -99)
na_if(x, -99)
```

-   Converts “special codes” into real `NA`.
-   Makes summaries and plots treat those values as missing correctly.

### `tidyr::complete(...)`

```{r}
library(tidyr)
library(dplyr)

df <- tibble(
  student = c("A","A","B"),
  week = c(1, 3, 1),
  score = c(10, 12, 9)
)

df |>
  complete(student, week = 1:3)
```

-   Creates rows for missing combinations of keys (implicit missing → explicit).
-   Helps you see gaps and then fill/plot time series cleanly.

### `anti_join(x, y)`

```{r}
library(dplyr)

x <- tibble(id = c(1,2,3), name = c("a","b","c"))
y <- tibble(id = c(2,3))

anti_join(x, y, by = "id")
```

-   Returns rows in `x` whose keys do **not** appear in `y`.
-   Useful for finding mismatches or missing records.

## Joins

### `left_join(x, y, by = ...)`

```{r}
library(dplyr)

students <- tibble(id = c(1,2,3), name = c("A","B","C"))
grades   <- tibble(id = c(1,3), grade = c(90, 85))

left_join(students, grades, by = "id")
```

-   Keeps all rows from `x` and adds matching columns from `y`.
-   Unmatched rows get `NA` for new columns.

### `inner_join()` / `full_join()` / `right_join()`

```{r}
library(dplyr)

inner_join(students, grades, by = "id")
full_join(students, grades, by = "id")
right_join(students, grades, by = "id")
```

-   `inner_join()` keeps only rows with matches in both tables.
-   `full_join()` keeps everything from both; `right_join()` keeps all rows of `y`.

### `semi_join()` / `anti_join()`

```{r}
library(dplyr)

semi_join(students, grades, by = "id")
anti_join(students, grades, by = "id")
```

-   `semi_join()` filters `x` to rows that have a match in `y` (no extra columns added).
-   `anti_join()` filters `x` to rows that do **not** have a match in `y`.
