---
title: "R Code Summary: Activities 11–19"
format:
  html:
    toc: true
    toc-depth: 2
editor: visual
execute:
  warning: false
  message: false
---

```{r setup}
# This setup chunk makes the document safer to run.
# It loads packages only if they are installed.

pkg_available <- function(pkg) {
  requireNamespace(pkg, quietly = TRUE)
}

load_pkg <- function(pkg) {
  if (pkg_available(pkg)) {
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
    TRUE
  } else {
    message("Package not installed: ", pkg)
    FALSE
  }
}

load_pkg("dplyr")
load_pkg("ggplot2")
load_pkg("purrr")
load_pkg("tibble")
load_pkg("stringr")
load_pkg("tidyr")
```

## Activity 11: Iteration I

### `for` loop over a vector

```{r}
groups <- c("group1", "group2", "group3")

for (i in seq_along(groups)) {
  print(groups[i])
}
```

Use `seq_along()` when looping through a vector.

### Store output from a loop

```{r}
char_storage <- vector("character", length = 3)
num_storage <- vector("numeric", length = 3)
list_storage <- vector("list", length = 3)

for (i in seq_len(3)) {
  char_storage[i] <- paste0("Number: ", i)
  num_storage[i] <- 2 * i
  list_storage[[i]] <- i
}

char_storage
num_storage
list_storage
```

Create storage before the loop when you want to save output.

### Loop over rows of a data frame

```{r}
small_data <- tibble::tibble(a = 1:2, b = 2:3, c = 4:5)

for (r in seq_len(nrow(small_data))) {
  print(small_data[r, ])
}
```

Use `seq_len(nrow(data))` when looping through rows.

### `across()` for multiple columns

```{r}
if (pkg_available("dplyr") && pkg_available("ggplot2")) {
  ggplot2::diamonds |>
    dplyr::mutate(
      dplyr::across(c(carat, x, y, z), ~ round(.x, digits = 1))
    ) |>
    dplyr::select(carat, x, y, z) |>
    head()
}
```

`across()` applies the same function to many columns.

### `across()` with summary functions

```{r}
if (pkg_available("dplyr") && pkg_available("ggplot2")) {
  ggplot2::diamonds |>
    dplyr::group_by(cut) |>
    dplyr::summarize(
      dplyr::across(
        c(carat, price),
        list(mean = ~ mean(.x, na.rm = TRUE), sd = ~ sd(.x, na.rm = TRUE)),
        .names = "{.col}_{.fn}"
      ),
      .groups = "drop"
    )
}
```

## Activity 12: Iteration II

### Anonymous functions

```{r}
x <- c(1, 2, 3, NA, 5)

median_no_na <- function(x) median(x, na.rm = TRUE)
median_no_na(x)

# Anonymous function version
(function(x) median(x, na.rm = TRUE))(x)
```

An anonymous function is a throw-away function with no saved name.

### `map()` returns a list

```{r}
if (pkg_available("purrr")) {
  numbers <- list(a = 1:3, b = 4:6, c = 7:9)
  purrr::map(numbers, mean)
}
```

Use `map()` when you want output saved.

### Typed map functions

```{r}
if (pkg_available("purrr")) {
  numbers <- list(a = 1:3, b = 4:6, c = 7:9)
  purrr::map_dbl(numbers, mean)
  purrr::map_chr(numbers, ~ paste(.x, collapse = ", "))
}
```

`map_dbl()`, `map_chr()`, and similar functions control the output type.

### `map2()` and `pmap()`

```{r}
if (pkg_available("purrr")) {
  purrr::map2_chr(
    c("A", "B", "C"),
    c(90, 85, 92),
    ~ paste(.x, "scored", .y)
  )

  student_scores <- tibble::tibble(
    name = c("A", "B", "C"),
    exam = c("Exam 1", "Exam 1", "Exam 1"),
    score = c(90, 85, 92)
  )

  purrr::pmap_chr(student_scores, ~ paste(..1, ..2, ..3))
}
```

`map()` uses one input, `map2()` uses two inputs, and `pmap()` uses many inputs.

### `walk()` for side effects

```{r}
if (pkg_available("purrr")) {
  purrr::walk(c("Hello", "from", "walk"), print)
}
```

Use `walk()` when you want an action, like printing or saving a file, but do not need returned output.

### Nested data and models

```{r}
if (pkg_available("dplyr") && pkg_available("tidyr") && pkg_available("purrr")) {
  mtcars_nested <- mtcars |>
    tibble::rownames_to_column("car") |>
    dplyr::group_by(cyl) |>
    tidyr::nest()

  mtcars_nested <- mtcars_nested |>
    dplyr::mutate(model = purrr::map(data, ~ lm(mpg ~ wt + hp, data = .x)))

  mtcars_nested
}
```

This fits a separate model for each group.

### `pluck()`, `keep()`, and `reduce()`

```{r}
if (pkg_available("purrr")) {
  nested_list <- list(
    a = list(value = 1),
    b = list(value = 2),
    c = list(value = 3)
  )

  purrr::pluck(nested_list, "b", "value")

  purrr::keep(1:10, ~ .x > 5)

  purrr::reduce(1:5, `+`)
}
```

`pluck()` extracts deeply nested elements, `keep()` filters lists/vectors, and `reduce()` combines many values into one.

## Activity 13: APIs I and JSON

### JSON object to R

```{r}
if (pkg_available("jsonlite")) {
  json_text <- '{"name": "Mario", "age": 32, "city": "New York"}'
  jsonlite::fromJSON(json_text)
}
```

JSON is a format for organizing data, often from APIs.

### JSON array of objects to data frame

```{r}
if (pkg_available("jsonlite")) {
  json_text <- '[
    {"name": "Mario", "age": 32},
    {"name": "Peach", "age": 21},
    {"name": "Luigi", "age": 30}
  ]'

  jsonlite::fromJSON(json_text)
}
```

A JSON array of similar objects often becomes a data frame in R.

### Nested JSON to named list

```{r}
if (pkg_available("jsonlite")) {
  json_text <- '{
    "result_count": 3,
    "results": [
      {"_href": "url1", "_schema": "schema1", "type": "LogicalSwitchConfig"},
      {"_href": "url2", "_schema": "schema2", "type": "LogicalSwitchConfig"},
      {"_href": "url3", "_schema": "schema3", "type": "LogicalSwitchConfig"}
    ]
  }'

  parsed <- jsonlite::fromJSON(json_text)
  parsed$result_count
  parsed$results
}
```

This structure makes sense as a named list with two elements: `result_count` and `results`.

## Activity 14: APIs II

### URL parts

```{r}
url <- "https://www.tutorialspoint.com/cgi-bin/search.cgi?searchTerm=HTML#top"

parts <- list(
  scheme = "https://",
  host = "www.tutorialspoint.com",
  path = "/cgi-bin/search.cgi",
  query = "searchTerm=HTML",
  fragment = "#top"
)

parts
```

URLs commonly contain a scheme, host, file path, query, and optional fragment identifier.

### URL encoding and decoding

```{r}
text <- "cats & dogs"
encoded <- utils::URLencode(text, reserved = TRUE)
decoded <- utils::URLdecode(encoded)

encoded
decoded
```

Encoding makes strings safe to use inside URLs.

### Build and read from a simple API-style URL

```{r}
# This example is not evaluated online. It shows the usual pattern.
base_url <- "https://api.example.com/data"
query <- list(searchTerm = "HTML", limit = 10)

api_url <- paste0(
  base_url,
  "?",
  paste(names(query), query, sep = "=", collapse = "&")
)

api_url
```

A real API request usually combines a base URL, path, and query parameters.

### `httr2` request pattern

```{r}
if (pkg_available("httr2")) {
  request_example <- httr2::request("https://api.example.com") |>
    httr2::req_url_path_append("data") |>
    httr2::req_url_query(searchTerm = "HTML", limit = 10)

  request_example
}
```

Use `httr2::request()`, `req_url_path_append()`, and `req_url_query()` to build API requests.

## Activity 15: Web scraping

### HTML structure

```{r}
html_example <- "<h1 id='first'>A heading</h1>"
html_example
```

In this example, `<h1 id='first'>` is the start tag, `id='first'` is the attribute, `A heading` is the content, and `</h1>` is the end tag.

### CSS selector examples with local HTML

```{r}
if (pkg_available("rvest")) {
  html <- rvest::read_html(
    "<html>
      <body>
        <h1 id='title'>This is a heading</h1>
        <p id='first'>This is a paragraph</p>
        <p class='important'>This is an important paragraph</p>
      </body>
    </html>"
  )

  # Select all paragraph elements
  html |>
    rvest::html_elements("p") |>
    rvest::html_text2()

  # Select elements with class = 'important'
  html |>
    rvest::html_elements(".important") |>
    rvest::html_text2()

  # Select element with id = 'title'
  html |>
    rvest::html_element("#title") |>
    rvest::html_text2()
}
```

`p` selects paragraph tags, `.important` selects a class, and `#title` selects an ID.

### Extract HTML attributes

```{r}
if (pkg_available("rvest")) {
  html <- rvest::read_html(
    "<html><body><a href='https://example.com'>Example link</a></body></html>"
  )

  html |>
    rvest::html_element("a") |>
    rvest::html_attr("href")
}
```

Use `html_attr()` to get attributes such as links.

## Activity 16: Databases and SQL

### Connect to a local DuckDB database

```{r}
if (pkg_available("DBI") && pkg_available("duckdb") && pkg_available("dplyr")) {
  con <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")

  DBI::dbWriteTable(con, "mtcars", mtcars, overwrite = TRUE)

  mtcars_db <- dplyr::tbl(con, "mtcars")
  mtcars_db

  DBI::dbDisconnect(con, shutdown = TRUE)
}
```

The first step for using a database in R is to establish a connection.

### Use dplyr on a database table

```{r}
if (pkg_available("DBI") && pkg_available("duckdb") && pkg_available("dplyr")) {
  con <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")
  DBI::dbWriteTable(con, "mtcars", mtcars, overwrite = TRUE)

  mtcars_db <- dplyr::tbl(con, "mtcars")

  big_cars <- mtcars_db |>
    dplyr::filter(hp > 100) |>
    dplyr::select(mpg, cyl, hp, wt)

  big_cars |>
    dplyr::show_query()

  big_cars |>
    dplyr::collect()

  DBI::dbDisconnect(con, shutdown = TRUE)
}
```

Use `collect()` only after filtering/selecting the data you need.

### Basic SQL query

```{r}
if (pkg_available("DBI") && pkg_available("duckdb")) {
  con <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")
  DBI::dbWriteTable(con, "mtcars", mtcars, overwrite = TRUE)

  DBI::dbGetQuery(con, "
    SELECT mpg, cyl, hp, wt
    FROM mtcars
    WHERE hp > 100
    ORDER BY mpg DESC
    LIMIT 5
  ")

  DBI::dbDisconnect(con, shutdown = TRUE)
}
```

SQL is used to query a subset of data from a database.

## Activity 17: Presenting visualizations

### Add clear labels

```{r}
if (pkg_available("ggplot2")) {
  ggplot2::ggplot(mtcars, ggplot2::aes(x = wt, y = mpg)) +
    ggplot2::geom_point() +
    ggplot2::labs(
      title = "Heavier cars tend to have lower gas mileage",
      subtitle = "Example using the built-in mtcars data",
      x = "Weight (1000 lbs)",
      y = "Miles per gallon",
      caption = "Data: mtcars"
    )
}
```

Good visualizations use clear titles, subtitles, axis labels, and captions.

### Use faceting to compare groups

```{r}
if (pkg_available("ggplot2")) {
  ggplot2::ggplot(mtcars, ggplot2::aes(x = wt, y = mpg)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~ cyl) +
    ggplot2::labs(
      title = "Weight and mileage by cylinder group",
      x = "Weight (1000 lbs)",
      y = "Miles per gallon"
    )
}
```

Facets split one plot into small multiples.

### Reorder categories before plotting

```{r}
if (pkg_available("dplyr") && pkg_available("ggplot2") && pkg_available("forcats")) {
  mtcars_summary <- mtcars |>
    tibble::rownames_to_column("car") |>
    dplyr::mutate(cyl = factor(cyl)) |>
    dplyr::group_by(cyl) |>
    dplyr::summarize(mean_mpg = mean(mpg), .groups = "drop") |>
    dplyr::mutate(cyl = forcats::fct_reorder(cyl, mean_mpg))

  ggplot2::ggplot(mtcars_summary, ggplot2::aes(x = cyl, y = mean_mpg)) +
    ggplot2::geom_col() +
    ggplot2::labs(x = "Cylinders", y = "Mean MPG")
}
```

Ordering categories can make visual comparisons easier.

## Activity 18: Interactive visualization

### Simple Shiny app code

```{r}
# This chunk prints Shiny app code instead of launching an app during rendering.
cat('
library(shiny)
library(ggplot2)

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("cyl", "Choose cylinders:", choices = unique(mtcars$cyl))
    ),
    mainPanel(
      plotOutput("scatterplot")
    )
  )
)

server <- function(input, output) {
  output$scatterplot <- renderPlot({
    mtcars |>
      dplyr::filter(cyl == input$cyl) |>
      ggplot(aes(x = wt, y = mpg)) +
      geom_point()
  })
}

shinyApp(ui = ui, server = server)
')
```

A Shiny app has a `ui` section and a `server` section.

### Plotly from ggplot

```{r}
if (pkg_available("ggplot2") && pkg_available("plotly")) {
  p <- ggplot2::ggplot(mtcars, ggplot2::aes(x = wt, y = mpg, color = factor(cyl))) +
    ggplot2::geom_point() +
    ggplot2::labs(color = "Cylinders")

  plotly::ggplotly(p)
}
```

`ggplotly()` turns many ggplot graphics into interactive graphics.

## Activity 19: Code and data quality

### DRY: do not repeat yourself

```{r}
if (pkg_available("dplyr")) {
  mtcars |>
    dplyr::summarize(
      dplyr::across(c(mpg, hp, wt), ~ mean(.x, na.rm = TRUE))
    )
}
```

Use functions or `across()` instead of copying and pasting similar code.

### Check missing values

```{r}
if (pkg_available("dplyr")) {
  example_data <- tibble::tibble(
    id = 1:4,
    score = c(90, NA, 85, NA),
    group = c("A", "A", NA, "B")
  )

  example_data |>
    dplyr::summarize(
      dplyr::across(dplyr::everything(), ~ sum(is.na(.x)))
    )
}
```

Always check missing values after reading or cleaning data.

### Check joins with `anti_join()`

```{r}
if (pkg_available("dplyr")) {
  students <- tibble::tibble(id = c(1, 2, 3), name = c("A", "B", "C"))
  grades <- tibble::tibble(id = c(1, 3), grade = c(90, 85))

  dplyr::anti_join(students, grades, by = "id")
}
```

Use `anti_join()` to find rows that did not match.

### Benchmark slow vs. fast code

```{r}
if (pkg_available("bench")) {
  bench::mark(
    slow = {
      x <- integer()
      for (i in 1:1000) {
        x <- c(x, i)
      }
    },
    fast = {
      x <- vector("integer", length = 1000)
      for (i in 1:1000) {
        x[i] <- i
      }
    },
    iterations = 5,
    check = FALSE
  )
}
```

Pre-allocating storage is usually faster than growing an object inside a loop.

## Quick reference

| Topic                        | Use this                 |
|------------------------------|--------------------------|
| Repeat with saved output     | `map()` or `map_*()`     |
| Repeat with side effect      | `walk()`                 |
| Repeat across columns        | `across()`               |
| Extract list element         | `[[ ]]` or `pluck()`     |
| Filter list/vector           | `keep()`                 |
| Combine many into one        | `reduce()`               |
| Parse JSON                   | `jsonlite::fromJSON()`   |
| Scrape HTML elements         | `rvest::html_elements()` |
| Extract HTML text            | `rvest::html_text2()`    |
| Extract HTML attribute       | `rvest::html_attr()`     |
| Connect to database          | `DBI::dbConnect()`       |
| Bring database result into R | `collect()`              |
| Make ggplot interactive      | `plotly::ggplotly()`     |
| Check unmatched rows         | `anti_join()`            |
