---
title: "Homework 02"
---


## TidyTuesday Section (optional)

::: {.callout-important title="Instructions"}

You can count work on this week's TidyTuesday toward the exceptional work required for an A in the Homework component.

:::


Explore the week's [TidyTuesday](https://github.com/rfordatascience/tidytuesday) challenge. Develop a research question, then answer it through a short data story with [effective visualization(s)]({{< var effective-viz-url >}}).  Provide sufficient background for readers to grasp your narrative.


## Research question

Have LEGO set prices increased faster than set size (piece count) over time, and what does that mean for “value” (price per piece)?

## Background

LEGO fans often compare sets by *price per piece* as a rough measure of value. Over time, sets have changed in complexity, licensing, and production costs. Using the LEGO Sets TidyTuesday data, I compare how piece counts and prices move across years, and whether price-per-piece has trended upward.

## Data

```{r}
library(tidyverse)
library(tidytuesdayR)
library(scales)

tt <- tt_load("2022-09-06")
sets <- tt$`sets.csv`
```



```{r}
sets_clean <- sets |>
  transmute(
    year = as.integer(year),
    name,
    theme_id,
    pieces = as.numeric(num_parts)
  )

```

```{r}
ggplot(sets_clean, aes(x = year, y = pieces)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm") +
  labs(x = "Year", y = "Number of parts")

```
```{r}
sets_clean <- sets_clean |>
  mutate(decade = floor(year / 10) * 10)

ggplot(sets_clean, aes(x = factor(decade), y = pieces)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_y_log10() +
  labs(
    x = "Decade",
    y = "Number of parts (log scale)",
    title = "LEGO set sizes have increased over time",
    subtitle = "Distribution of number of parts by decade"
  ) +
  theme_minimal()

```

This analysis uses LEGO set data from TidyTuesday to examine how LEGO sets have changed over time.
My research question asks whether LEGO sets have become larger, as measured by the number of pieces, across different years.
The first plot shows the relationship between release year and number of pieces, with a fitted trend line indicating that newer LEGO sets generally contain more pieces than older ones.
While there is substantial variation in set size within each year, the overall upward trend suggests increasing complexity in LEGO sets over time.
The second plot compares themes by their median number of pieces, focusing on themes with a sufficient number of sets.
This comparison shows that some themes consistently feature larger sets, highlighting how theme design influences set size beyond simple time trends.
